---
title: "Fitting rainfall profiles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting-rainfall-profiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(umbrella)
```

Below is an example workflow for using the umbrella package to fit rainfall profiles
for use in malariasimulation.

First you'll need to download the rasters and extract the raw data. There are 
some helper function to download global daily rainfall rasters in the package.
You can use the `get_urls()` function to create the URLs to get daily rainfall
rasters for a given year. After that the `download_raster()` can be used to 
download the rasters to your device. Beware that the set of unzipped rasters
for a whole year will take up quite a bit of memory (each raster is about 2.2 MB).

Once you have you set of rasters you can load them and extract the data. For this
I recommend getting a `sf` shape file with the boundaries of the area you
would like to extract the data for. The R package `terra` is a great way to
extract the data, something like:

```{r, eval=FALSE}
my_raster_files <- c("rainfall1.tif", "rainfall2.tif")
my_raster <- terra::rast(my_raster_files)
my_sf <- load("my_sf.rds")
extract_data <- terra::extract(my_raster, my_sf, fun = mean)
```

Once you've extracted and tidied the output, you might have somethig like:

```{r}
rainfall
```

Lets fit a simple seasonal profile to this dataset, and a profile with an imposed
minimum rainfall floor.

```{r, fig.width=7}
fit1 <- fit_fourier(rainfall = rainfall$rainfall, t = rainfall$day, floor = 0)
predict_1 <- fourier_predict(coef = fit1$coefficients, t = 1:365, floor = 0)

fit2 <- fit_fourier(rainfall = rainfall$rainfall, t = rainfall$day, floor = 1)
predict_2 <- fourier_predict(coef = fit1$coefficients, t = 1:365, floor = 1)

plot(rainfall, xlab = "Day", ylab = "Rainfall", pch = 19)
lines(predict_1, col = "deeppink", lwd = 2)
lines(predict_2, col = "dodgerblue", lwd = 2, lty = 2)
```
