---
title: "Fitting rainfall profiles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting-rainfall-profiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(umbrella)
```

Below is an example workflow for using the umbrella package to fit rainfall profiles
for use in malariasimulation. Fitting a Fourier series provides a smooth
representation of seasonal rainfall patterns that can be supplied directly to the
`malariasimulation` package.

First you'll need to download the rasters and extract the raw data. There are
some helper functions to download global daily rainfall rasters in the package.
You can use the `get_urls()` function to create the URLs to get daily rainfall
rasters for a given year. After that the `download_raster()` can be used to 
download the rasters to your device. Beware that the set of unzipped rasters
for a whole year will take up quite a bit of memory (each raster is about 2.2 MB).

Once you have your set of rasters you can load them and extract the data. For this
I recommend getting a `sf` shape file with the boundaries of the area you
would like to extract the data for. The R package `terra` is a great way to
extract the data, something like:
Many openly available administrative boundary shapefiles can be obtained from
resources such as [GADM](https://gadm.org/) or [Natural Earth](https://www.naturalearthdata.com/).

```{r, eval=FALSE}
my_raster_files <- c("rainfall1.tif", "rainfall2.tif")
my_raster <- terra::rast(my_raster_files)
my_sf <- load("my_sf.rds")
extract_data <- terra::extract(my_raster, my_sf, fun = mean)
```

Once you've extracted and tidied the output, you might have something like:

```{r}
head(rainfall)
```

The built-in `rainfall` data set contains 365 rows with two columns:
`day`, giving the day of the year, and `rainfall`, giving the observed
rainfall in millimetres at a single location. We'll use it here to
demonstrate fitting seasonal profiles.

Let's fit a simple seasonal profile to this dataset, and a profile with an imposed
minimum rainfall floor.

```{r, fig.width=5, fig.height=5}
fit1 <- fit_fourier(rainfall = rainfall$rainfall, t = rainfall$day, floor = 0)
predict_1 <- fourier_predict(coef = fit1$coefficients, t = 1:365, floor = 0)

fit2 <- fit_fourier(rainfall = rainfall$rainfall, t = rainfall$day, floor = 1)
predict_2 <- fourier_predict(coef = fit2$coefficients, t = 1:365, floor = 1)

plot(rainfall, xlab = "Day", ylab = "Rainfall", pch = 19)
lines(predict_1, col = "deeppink", lwd = 2)
lines(predict_2, col = "dodgerblue", lwd = 2, lty = 2)
```

The pink line shows the profile fitted without imposing a minimum rainfall
value. The dashed blue line applies a floor of 1 mm and therefore never drops
below that value.
