---
title: "Fitting-rainfall-profiles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting-rainfall-profiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(umbrella)
library(sf)
library(dplyr)
library(tidyr)
library(ggplot2)
```

Below is an example workflow for using the umbrella package to fit rainfall profiles
for use in malariasimulation.

First we need a sf shape file to define the areas we want to capture rainfall for:

```{r}
# Load an example sf file for two admin 1 units
spatial_data <- umbrella::example_admin_sf
```

Specify the start and end times of the period we are interested in

```{r}
# Specify time-range
start_date <- "2017-01-01"
end_date <- "2019-12-31"
```

And extract the data (We won't do this here as it links to python and takes a little time)

```{r, eval=FALSE}
# Extract
daily_rain_raw <- pull_daily_rainfall(sf = spatial_data, start_date = start_date, end_date = end_date)
```

Instead we can load pre-extracted output

```{r}
# Extract
daily_rain_raw <- umbrella::example_daily_rain_raw
```



Next, let's tidy up the raw output

```{r}
# Process the raw data
daily_rain  <- daily_rain_raw %>%
  # Convert to long and format
  pivot_longer(-c(ISO, Country, admin1),
               names_to = "date",
               values_to = "rainfall",
               names_prefix = "X",
               names_pattern = "(.*)_precipitation") %>%
  mutate(date = as.Date(as.character(readr::parse_number(.data$date)), format = "%Y%m%d"),
         year = lubridate::year(.data$date),
         day_of_year = lubridate::yday(.data$date),
         t = lubridate::yday(.data$date) / 365,
         rainfall = as.numeric(rainfall)) %>%
  # Replace missing data with 0
  replace_na(replace = list(rainfall = 0)) %>%
  # Remove any leap year addtional days
  dplyr::filter(.data$day_of_year < 366)
```

First we will fit a profile combining data from all three years:

```{r, fig.width=7}
rain <- daily_rain %>%
  group_by(ISO, Country, admin1) %>%
  summarise(
    data = list(data.frame(date, day_of_year, t, rainfall)),
    model = list(fit_fourier(data[[1]]$rainfall, data[[1]]$t)),
    profile = list(fourier_predict(model[[1]]$coefficients, t = 1:365 / 365, floor = model[[1]]$floor))
  )

# Visualise output
rd <- rain %>%
  select(ISO, Country, data) %>%
  tidyr::unnest(cols = c("data"))
pf <- rain %>%
  select(ISO, Country, profile) %>%
  tidyr::unnest(cols = c("profile"))

ggplot() +
  geom_point(data = rd, aes(x = t, y = rainfall, col = Country), alpha = 1, size = 0.1) +
  geom_line(data = pf, aes(x = t, y = profile)) +
  theme_bw() +
  theme(strip.background = element_rect(fill = "white")) +
  facet_grid( ~ Country)
```

If required we can fit each year individually:

```{r, fig.width=7, fig.height=7}
rain2 <- daily_rain %>%
  group_by(year, ISO, Country, admin1) %>%
  summarise(
    data = list(data.frame(date, day_of_year, t, rainfall)),
    model = list(fit_fourier(data[[1]]$rainfall, data[[1]]$t)),
    profile = list(fourier_predict(model[[1]]$coefficients, t = 1:365 / 365, floor = model[[1]]$floor))
  )

# Visualise output
rd2 <- rain2 %>%
  select(year, ISO, Country, data) %>%
  tidyr::unnest(cols = c("data"))
pf2 <- rain2 %>%
  select(year, ISO, Country, profile) %>%
  tidyr::unnest(cols = c("profile"))

ggplot() +
  geom_point(data = rd2, aes(x = t, y = rainfall, col = Country), alpha = 1, size = 0.1) +
  geom_line(data = pf2, aes(x = t, y = profile)) +
  theme_bw() +
  theme(strip.background = element_rect(fill = "white")) +
  facet_grid(year ~ Country)

```
